### 2021-02-09 final_report/include/CE.tex
quickly reviewed CE chapter

RTree {
 root = UDWord {
  udID = 3, udFORM = "tree", udLEMMA = "tree", udUPOS = "NOUN", (...), udHEAD = 0,
  udDEPREL = "root", (...)}, 
 subtrees = [
  RTree {
  root = UDWord {
   udID = 1, udFORM = "another", udLEMMA = "another", udUPOS = "DET", (...),
   udHEAD = 3, udDEPREL = "det", (...)}, 
  subtrees = []},
  RTree {
  root = UDWord {
   udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
   udHEAD = 3, udDEPREL = "compound", (...)}, 
  subtrees = []}]}

@@

 \item $\langle$\textit{we are perfectly aligned, siamo perfettamente allineati}$\rangle$
 \item $\langle$\textit{aligned,allineati}$\rangle$ (head alignment)
 \item $\langle$\textit{are}$\rangle$
 \item $\langle$\textit{siamo}$\rangle$
 \item $\langle$\textit{perfectly,perfettamente}$\rangle$
 \item $\langle$\textit{aligned,allineati}$\rangle$

@@

There are indeed other divergences of this kind that can occur in a parallel text and cannot be captured by Criterion \ref{udmatch}, such as
\begin{example}
 ``Claudio is \underline{hungry}'' VS ``Claudio ha \underline{fame}''
\end{example}
where, in the original sentence, the complement of the copula \textit{am}, \textit{hungry}, would be labelled as its \texttt{root} but, in Italian, the noun \textit{fame} is the object of the root verb \textit{ha}. 
It proved hard to detect cases like this via purely syntactic rules without causing the program to extract a lot of incorrect alignments as well, but if necessary it is easy to modify the program by removing or adding rules of this kind, potentially even language-specific.

\subparagraph*{Conflational divergence} \label{confl}
Conflational divergences involve the translation of two or more words in the source language using a single word that combines their meanings in the target language. A straightforward example is that of compounds, for instance
\begin{example} 
 ``Filippo is interested in \underline{game design}'' VS ``Filippo är intresserad av \underline{spelutvekling}''
\end{example}
but there are also other cases, like
\begin{example} 
 ``I'll come and \underline{say hello} to Bruno and Andrea'' VS ``Verrò a \underline{salutare} Bruno e Andrea''
\end{example} 
where the single word \textit{salutare} is not a compound but expresses the same meaning as ``say'' and ``hi'' together. \smallskip
Divergences like the one described in the latter example are usually captured by Criterion \ref{udmatch}, while compounds often need to be taken care of explicitly. For reasons that will become clear later, these cases are discussed in Section \ref{heads}.

\subparagraph*{Structural divergence}
Structural divergences happen when the subject, object or indirect object of a sentence in the source language are rendered as obliques in the source language. For instance, in the sentences
\begin{example}
 ``I called \underline{Francesco}'' VS ``Ho telefonato \underline{a Francesco}''
\end{example}
what is the direct object in English (\textit{Francesco}) becomes a prepositional phrase (\textit{a Francesco}) in Italian. 
Even though in many cases these divergences can be captured by POS-equivalence, in order to ensure higher precision, the CE module handles all of them explicitly via rules that combine it with UD label checking and whose priority is higher than that of the basic Criterion \ref{poseq}.

\subparagraph*{Head swapping divergences}
Head swapping divergences always involve a head verb and a logical modifier and occur when the logical modifier is placed lower (resp. higher) in the source language sentence than in its target language counterpart. For instance, in the following example
\begin{example}
 ``Anna \underline{usually} goes for walks'' VS ``Anna \underline{brukar} promenera''
\end{example} 
in Swedish, the logical modifier \textit{usually} is implicitly expressed by the verb \textit{brukar} itself, i.e. placed ``higher up''. \smallskip

@@

One could object that divergences like the above (and some other) could - and maybe \textit{should} - be avoided when translating a text. We will not dwell upon this, since what matters for the purpose of CA is not what is desirable in natural language translation but rather what divergences do occur in the human-translated parallel texts available for analysis. \smallskip

@@

\paragraph{Pruning alignments} \label{pruning} 
In its most basic version, this algorithm does nothing to avoid subtrees in the source language to be aligned with multiple subtrees in the target language and vice versa. As a consequence, it systematically \textit{overgenerates} alignments. \smallskip

@@

It can happen, however, that two trees match \textit{more than one} of the criteria. For instance, two trees might have their roots sharing the same UD label (cf. Criterion \ref{udmatch}) \textit{and} be POS-equivalent. In such cases, it intuitively makes sense to consider them more likely to be exact, as there are literally more reasons to align them. This makes the implicit ordering that the solution just described produces ``for free'' insufficient and leads to the necessity to keep track of the set of reasons why an alignment has been extracted and only then sort the alternative alignments. From the implementative point of view, each alignment is simply associated with a \texttt{Set} of labels, whose type is in fact named \texttt{Reason} and has an ordering defined over. The final sorting is based on:\smallskip

@@

 Without head alignments, the output of the algorithm described so far would be:
 \begin{itemize}
 \item $\langle$Enrico eats a banana, Enrico mangia una banana$\rangle$
 \item $\langle$Enrico, Enrico$\rangle$
 \item $\langle$a banana, una banana$\rangle$
 \item $\langle$a, una$\rangle$
 \end{itemize} 
 while introducing head alignment leads to detecting two additional, equally relevant one-word correspondences\footnote{note how, even with head alignments, not all potentially interesting alignments are found. In this example, for instance, it could be useful to align the verb phrase ``\textit{eats a banana}'' with its Italian counterpart ``\textit{mangia una banana}'', whose members are also not, strictly speaking, subtrees of the original sentences. The approach we use for head alignment, as we will discuss in Chapter \ref{ch6}, could be generalized to cases like this.}:
 \begin{itemize}
 \item $\langle$eats, mangia$\rangle$
 \item $\langle$banana, banana$\rangle$
 \end{itemize} 
\end{example}

While it is of extreme importance not to miss alignments like the last two in the example above, aligning heads is not always appropriate, especially when, as in the present case, common translation divergence patterns are one of the criteria. For instance, when two trees are aligned because of a categorial divergence such as that of Example \ref{herb}, where \textit{doctoral} corresponds to \textit{di dottorato}, it is at least questionable to also align \textit{doctoral} with \textit{dottorato}. Consequently, in the current implementation of the algorithm, each alignment criterion is associated with a flag telling whether heads should also be aligned whenever a new alignment is extracted because of that criterion. \smallskip

@@

\subparagraph*{Compounds}
We mentioned compounds as a type of conflational divergence (cf. Section \ref{confl}). Given the variety of ways in which the translation equivalent of a compound can be expressed and the subsequent variety of related UD labels (a compound can be rendered as another compound, as a \texttt{compound} (space-separated) expression, as a noun modified by an adjective and/or another noun, as a \texttt{flat} multiword expression, as a combination of these things...), compounds are slightly harder to align than main verb + auxiliaries constructions in practice, but the solution is conceptually the same: before aligning two heads, we check whether the list of their immediate dependents contains anything labelled \texttt{compound}, \texttt{flat}, \texttt{amod} or \texttt{nmod}. If that's the case in, for instance, only the sentence in the source language, then it is likely that the head of the tree in the target language is in fact a compound. If so, it is aligned not just with the head of the tree in the source language, but with the tree composed of the head and the list of subtrees labelled as \texttt{compound}, \texttt{flat}, \texttt{amod} or \texttt{nmod}. In this way, our program is generally able to find the counterparts of a compound even when they are very complex, such as in the following case:
\begin{example} 
 ``I took a course on \underline{Machine Learning techniques}'' VS ``Jag deltog i en kurs om \underline{maskininlärningstekniker}''
\end{example}

@@

it is hard to deny that \textit{boat} has been ``translated'' as \textit{treno}, ``train'', so our CE module would - and should - be able to detect it, but in most situations it is by no means desirable that such correspondence at the word level is made use of, for instance by storing it in an bilingual lexicon!
As a consequence, each alignment can be marked as:
\begin{itemize}
 \item correct and useful for translation (\texttt{+})\footnote{at least to some extent: we observe that perfect translation equivalents that can be replaced with each other in all contexts are extremely rare, just like perfect synonyms.}
 \item correct but not useful for translation (\texttt{=}). This means that the CE module is working as expected and that we are are faced with a divergence in the broader sense of the term, i.e. potentially due to an idiomatic usage of language
 \item incorrect (\texttt{-}).
\end{itemize} \smallskip

@@

With alignments in this format available, a simple way to evaluate a specific version of the CE module is to manually assess the correctness of each of the correspondences the program identifies and compute some statistics about them. 
A separate Haskell module, \texttt{EvAlign}, was written to parse files in the above format and compute a series of useful statistics: \smallskip
\begin{itemize}
 \item the total number of alignments found
 \item the number of \textit{distinct} alignments, i.e. the number of alignments with distinct linearizations
 \item the percentage of correct alignments
 \item the percentage of correct alignments that are also useful for translation
 \item the percentage of alignments found due to each combination of criteria, and how many of them are correct.
\end{itemize} \smallskip

@@

% Please add the following required packages to your document preamble:
% \usepackage[normalem]{ulem}
% \useunder{\uline}{\ul}{}
\begin{table}[H]
 \centering
 \small
 \begin{tabular}{|l|l|l|l|l|}
 \hline
   & \multicolumn{2}{l|}{\textbf{baseline}} & \multicolumn{2}{l|}{\textbf{improved version}} \\ \hline
 \textbf{}   & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv} \\ \hline
 \textbf{distinct alignments} & 1097  & 1257 & 1198   & 1314  \\ \hline
 \textbf{correct (+ and =)} & 830 (58.12\%) & 995 (79.15\%)  & 964 (80.46\%) & 1105 (84.03\%) \\ \hline
 \textbf{correct and useful (+)} & 776 (54.34\%) & 976 (77.64\%)  & 896 (74.79\%) & 1082 (82.28\%) \\ \hline
 \end{tabular}
 \caption[Comparison between the baseline and the improved version of the CE on manually annotated data]{Comparison between the baseline and the improved version of the CE module using manually annotated sentences from the PUD treebank in two different language pairs, English-Italian and English-Swedish.}
 \label{tbase}
 \end{table}

@@

\begin{table}[H]
 \small
 \centering
 \begin{tabular}{|l|l|l|l|l|}
 \hline
 \textbf{criteria}       & \multicolumn{2}{l|}{\textbf{alignments extracted}} & \multicolumn{2}{l|}{\textbf{correct (+ and =)}} \\ \hline
 \textbf{}       & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv}  \\ \hline
 \textbf{matching UD labels}      & 39.39\%  & 37.71\%  & 78.60\%  & 80.24\%  \\ \hline
 \textbf{POS-equivalence}       & 3\%  & 1.82\%  & 83.33\%  & 83.33\%  \\ \hline
 \textbf{known divergence}       & 2.08\%   & 1.36\%  & 52\%   & 55.55\%  \\ \hline
 \textbf{known alignment}       & 0\%  & 0\%  & -  & -  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\end{tabular}}  & 20.95\%  & 24.71\%  & 93.03\%  & 94.76\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + head alignment\end{tabular}}  & 21.78\%  & 19.61\%  & 67.81\%  & 76.74\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}POS-equivalence\\ + head alignment\end{tabular}}  & 1.33\%   & 1.29\%  & 75\%   & 64.70\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + known alignment\end{tabular}}  & 0.25\%   & 0.45\%  & 66.66\%  & 83.33\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\\ + head alignment\end{tabular}} & 10.85\%  & 12.69\%  & 95.38\%  & 90.41\%  \\ \hline
 \end{tabular}
 \caption[Criterion-specific statistics]{Criterion-specific statistics. The leftmost column specifies a criterion or combination of criteria. The central column indicates the percentage of alignments extracted because of each set of criteria, and the rightmost one specifies how many of them were marked as \texttt{+} or \texttt{-}.}
 \label{treas}
 \end{table}

\subsection{Working with raw text}
Of course, it is also interesting to compare the performance of the CE module on manually annotated data with the results obtained on the automatically parsed course plans corpora, summarized in Table \ref{tcereal}.
% Please add the following required packages to your document preamble:
% \usepackage[normalem]{ulem}
% \useunder{\uline}{\ul}{}
\begin{table}[H]
 \small
 \centering 
 \begin{tabular}{|l|l|l|}
 \hline
 \textbf{}   & \textbf{DMI (en-it, 798 sentences)} & \textbf{CSE (en-sv, 498 sentences)} \\ \hline
 \textbf{distinct alignments} & 352  & 529  \\ \hline
 \textbf{correct (+ and =)} & 243 (69.03\%) & 368 (69.56\%) \\ \hline
 \textbf{correct and useful (+)} & 229 (65.05\%) & 351 (66.35\%) \\ \hline
 \end{tabular}
 \caption[Performance of CE on barely sentence-aligned data]{Performance of CE on barely sentence-aligned data, parsed automatically. Two distinct corpora of different size have been used for the two different language pairs, so the two columns are not to be compared with each other, but rather with Table \ref{tbase}.}
 \label{tcereal}
\end{table}

@@

A Haskell script converts the CoNNL-U files that are fed to \texttt{AlignTrees} into the input format required by \texttt{fast\_align}, so that the exact same tokenization is used. \smallskip

@@

Because \texttt{fast\_align} is a purely statistical approach, we trained it not only on the same 100-sentence subset of PUD used for evaluating \texttt{AlignTrees} against the baseline, but also on the full 1000-sentences dataset. In the latter experiment, the alignments obtained for the last 900 sentences were discarded before evaluating the results.
Due to the time required to manually annotate the numerous alignments extracted by \texttt{fast\_align}, this evaluation has been conducted only for the most challenging language pair, English-Italian. 
For a fairer comparison, since \texttt{fast\_align} works at the word level, only the one-to-one, one-to-many and many-to-one word alignments produced by the CE module were kept.

\begin{table}[H]
 \centering
 \small
 \begin{tabular}{|l|l|l|l|}
 \hline
  & \textbf{\texttt{AlignTrees}} & \textbf{\texttt{fast\_align} 100} & \textbf{\texttt{fast\_align} 1000}\\ \hline
  \textbf{distinct alignments} & 716 & 1440 & 1435 \\ \hline
  \textbf{correct} & 536 (74.86\%) & 410 (28.47\%) & 656 (45.71\%)\\ \hline
  \textbf{correct and useful} & 491 (68.57\%) & 371 (25.76\%) & 590 (41.11\%)\\ \hline
 \end{tabular}
 \caption[Comparison between the improved CE module and \texttt{fast\_align}]{Comparison between the improved CE module and \texttt{fast\_align} trained on the first 100 only (column 2) and on the whole PUD corpus (column 3).}
 \label{tfast}
\end{table}

>>>>

 RTree {
  root = UDWord {
   udID = 3, udFORM = "tree", udLEMMA = "tree", udUPOS = "NOUN", (...), udHEAD = 0,
   udDEPREL = "root", (...)}, 
  udDEPREL = "root", (...)}, 
   udDEPREL = "root", (...)}, 
  subtrees = [
   RTree {
   root = UDWord {
    udID = 1, udFORM = "another", udLEMMA = "another", udUPOS = "DET", (...),
    udHEAD = 3, udDEPREL = "det", (...)}, 
   udHEAD = 3, udDEPREL = "det", (...)}, 
    udHEAD = 3, udDEPREL = "det", (...)}, 
   subtrees = []},
   RTree {
   root = UDWord {
    udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
   udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
    udID = 2, udFORM = "dependency", udLEMMA = "dependency", udUPOS = "NOUN", (...), 
    udHEAD = 3, udDEPREL = "compound", (...)}, 
   udHEAD = 3, udDEPREL = "compound", (...)}, 
    udHEAD = 3, udDEPREL = "compound", (...)}, 
   subtrees = []}]}

@@

 \item $\langle$\textit{``we are perfectly aligned'', ``siamo perfettamente allineati''}$\rangle$
 \item $\langle$\textit{``aligned'', ``allineati''}$\rangle$ (head alignment)
 \item $\langle$\textit{``are'', ``siamo''}$\rangle$
 \item $\langle$\textit{``perfectly'', ``perfettamente}$\rangle$
 \item $\langle$\textit{``aligned'', ``allineati}$\rangle$

@@

There are indeed other divergences of this kind that can occur in a parallel text and cannot be captured by Criterion \ref{udmatch}, such as
\begin{example}
 ``Claudio is \underline{hungry}'' VS ``Claudio ha \underline{fame}''
\end{example}
where, in the original sentence, the complement of the copula ``\textit{am}'', ``\textit{hungry}'', would be labelled as its \texttt{root} but, in Italian, the noun ``\textit{fame}'' is the object of the root verb ``\textit{ha}''. 
It proved hard to detect cases like this via purely syntactic rules without causing the program to extract a lot of incorrect alignments as well, but if necessary it is easy to modify the program by removing or adding rules of this kind, potentially even language-specific.

\subparagraph*{Conflational divergence} \label{confl}
Conflational divergences involve the translation of two or more words in the source language using a single word that combines their meanings in the target language. A straightforward example is that of compounds, for instance
\begin{example} 
 ``Filippo is interested in \underline{game design}'' VS ``Filippo är intresserad av \underline{spelutvekling}''
\end{example}
but there are also other cases, like
\begin{example} 
 ``I'll come and \underline{say hello} to Bruno and Andrea'' VS ``Verrò a \underline{salutare} Bruno e Andrea''
\end{example} 
where the single word ``\textit{salutare}'' is not a compound but expresses the same meaning as ``\textit{say}'' and ``\textit{hi}'' together. \smallskip
Divergences like the one described in the latter example are usually captured by Criterion \ref{udmatch}, while compounds often need to be taken care of explicitly. For reasons that will become clear later, these cases are discussed in Section \ref{heads}.

\subparagraph*{Structural divergence}
Structural divergences happen when the subject, object or indirect object of a sentence in the source language are rendered as obliques in the source language. For instance, in the sentences
\begin{example}
 ``I called \underline{Francesco}'' VS ``Ho telefonato \underline{a Francesco}''
\end{example}
what is the direct object in English (``\textit{Francesco}'') becomes a prepositional phrase (``\textit{a Francesco}'') in Italian. 
Even though in many cases these divergences can be captured by POS-equivalence, in order to ensure higher precision, the CE module handles all of them explicitly via rules that combine it with UD label checking and whose priority is higher than that of the basic Criterion \ref{poseq}.

\subparagraph*{Head swapping divergences}
Head swapping divergences always involve a head verb and a logical modifier and occur when the logical modifier is placed lower (resp. higher) in the source language sentence than in its target language counterpart. For instance, in the following example
\begin{example}
 ``Anna \underline{usually} goes for walks'' VS ``Anna \underline{brukar} promenera''
\end{example} 
in Swedish, the logical modifier ``\textit{usually}'' is implicitly expressed by the verb ``\textit{brukar}'' itself, i.e. placed ``higher up''. \smallskip

@@

One could object that divergences like the above (and some other) could - and maybe should - be avoided when translating a text. We will not dwell upon this, since what matters for the purpose of CA is not what is desirable in natural language translation but rather what divergences do occur in the human-translated parallel texts available for analysis. \smallskip

@@

\paragraph{Pruning alignments} \label{pruning} 
In its most basic version, this algorithm does nothing to avoid subtrees in the source language to be aligned with multiple subtrees in the target language and vice versa. As a consequence, it systematically overgenerates alignments. \smallskip

@@

It can happen, however, that two trees match more than one of the criteria. For instance, two trees might have their roots sharing the same UD label (cf. Criterion \ref{udmatch}) and be POS-equivalent. In such cases, it intuitively makes sense to consider them more likely to be exact, as there are literally more reasons to align them. This makes the implicit ordering that the solution just described produces ``for free'' insufficient and leads to the necessity to keep track of the set of reasons why an alignment has been extracted and only then sort the alternative alignments. From the implementative point of view, each alignment is simply associated with a \texttt{Set} of labels, whose type is in fact named \texttt{Reason} and has an ordering defined over. The final sorting is based on:\smallskip

@@

 Without head alignments, the output of the algorithm described so far would be:
 \begin{itemize}
 \item $\langle$``Enrico eats a banana'', ``Enrico mangia una banana''$\rangle$
 \item $\langle$``Enrico'', ``Enrico''$\rangle$
 \item $\langle$``a banana'', ``una banana''$\rangle$
 \item $\langle$``a'', ``una''$\rangle$
 \end{itemize} 
 while introducing head alignment leads to detecting two additional, equally relevant one-word correspondences\footnote{note how, even with head alignments, not all potentially interesting alignments are found. In this example, for instance, it could be useful to align the verb phrase ``\textit{eats a banana}'' with its Italian counterpart ``\textit{mangia una banana}'', whose members are also not, strictly speaking, subtrees of the original sentences. The approach we use for head alignment, as we will discuss in Chapter \ref{ch6}, could be generalized to cases like this.}:
 \begin{itemize}
 \item $\langle$``eats'', ``mangia''$\rangle$
 \item $\langle$``banana'', ``banana''$\rangle$
 \end{itemize} 
\end{example}

While it is of extreme importance not to miss alignments like the last two in the example above, aligning heads is not always appropriate, especially when, as in the present case, common translation divergence patterns are one of the criteria. For instance, when two trees are aligned because of a categorial divergence such as that of Example \ref{herb}, where ``\textit{doctoral}'' corresponds to ``\textit{di dottorato}'', it is at least questionable to also align ``\textit{doctoral}'' with ``\textit{dottorato}''. Consequently, in the current implementation of the algorithm, each alignment criterion is associated with a flag telling whether heads should also be aligned whenever a new alignment is extracted because of that criterion. \smallskip

@@

\subparagraph*{Compounds}
We mentioned compounds as a type of conflational divergence (cf. Section \ref{confl}). Given the variety of ways in which the translation equivalent of a compound can be expressed and the subsequent variety of related UD labels (a compound can be rendered as another compound, as a \texttt{compound} (space-separated) expression, as a noun modified by an adjective and/or another noun, as a \texttt{flat} multiword expression, as a combination of these things...), compounds are slightly harder to align than main verb + auxiliaries constructions in practice, but the solution is conceptually the same: before aligning two heads, we check whether the list of their immediate dependents contains anything labelled \texttt{compound}, \texttt{flat}, \texttt{amod} or \texttt{nmod}. If that's the case in, for instance, only the sentence in the source language, then it is likely that the head of the tree in the target language is in fact a compound. If so, it is aligned not just with the head of the tree in the source language, but with the tree composed of the head and the list of subtrees labelled as \texttt{compound}, \texttt{flat}, \texttt{amod} or \texttt{nmod}. In this way, our program is generally able to find the counterparts of a compound even when they are very complex, such as in the following case:
\begin{example} 
 ``Giorgio took a course on \underline{Machine Learning techniques}'' VS ``Giorgio deltog i en kurs om \underline{maskininlärningstekniker}''
\end{example}

@@

it is hard to deny that ``\textit{boat}'' has been ``translated'' as ``\textit{treno}'', ``\textit{train}'', so our CE module would - and should - be able to detect it, but in most situations it is by no means desirable that such correspondence at the word level is made use of, for instance by storing it in an bilingual lexicon!
As a consequence, each alignment can be marked as:
\begin{itemize}
 \item correct and useful for translation (\texttt{+})\footnote{at least to some extent: we observe that perfect translation equivalents that can be replaced with each other in all contexts are extremely rare, just like perfect synonyms.}
 \item correct but not useful for translation (\texttt{=}). This means that the CE module is working as expected and that we are are faced with a divergence in the broader sense of the term, i.e. potentially due to an idiomatic usage of language
 \item incorrect (\texttt{-}).
\end{itemize} \smallskip

@@

With alignments in this format available, a simple way to evaluate a specific version of the CE module is to manually assess the correctness of each of the correspondences the program identifies and compute some statistics about them. 
A separate Haskell module, whose usage is described in Appendix \ref{b}, was written to parse files in the above format and compute a series of useful statistics: \smallskip
\begin{itemize}
 \item the total number of alignments found
 \item the number of distinct alignments, i.e. the number of alignments with distinct linearizations
 \item the percentage of correct alignments
 \item the percentage of correct alignments that are also useful for translation
 \item the percentage of alignments found due to each combination of criteria, and how many of them are correct.
\end{itemize} \smallskip

@@

% Please add the following required packages to your document preamble:
% \usepackage[normalem]{ulem}
% \useunder{\uline}{\ul}{}
\begin{table}[h]
 \centering
 \footnotesize
 \begin{tabular}{|l|l|l|l|l|}
 \hline
   & \multicolumn{2}{l|}{\textbf{baseline}} & \multicolumn{2}{l|}{\textbf{improved version}} \\ \hline
 \textbf{}   & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv} \\ \hline
 \textbf{distinct alignments} & 1097  & 1257 & 1198   & 1314  \\ \hline
 \textbf{correct (+ and =)} & 830 (58.12\%) & 995 (79.15\%)  & 964 (80.46\%) & 1105 (84.03\%) \\ \hline
 \textbf{correct and useful (+)} & 776 (54.34\%) & 976 (77.64\%)  & 896 (74.79\%) & 1082 (82.28\%) \\ \hline
 \end{tabular}
 \caption[Comparison between the baseline and the improved version of the CE on manually annotated data]{Comparison between the baseline and the improved version of the CE module using manually annotated sentences from the PUD treebank in two different language pairs, English-Italian and English-Swedish.}
 \label{tbase}
 \end{table}

@@

\begin{table}[H]
 \footnotesize
 \centering
 \begin{tabular}{|l|l|l|l|l|}
 \hline
 \textbf{criteria}       & \multicolumn{2}{l|}{\textbf{alignments extracted}} & \multicolumn{2}{l|}{\textbf{correct (+ and =)}} \\ \hline
 \textbf{}       & \textbf{en-it}  & \textbf{en-sv} & \textbf{en-it}  & \textbf{en-sv}  \\ \hline
 \textbf{matching UD labels}      & 39.39\%  & 37.71\%  & 78.60\%  & 80.24\%  \\ \hline
 \textbf{POS-equivalence}       & 3\%  & 1.82\%  & 83.33\%  & 83.33\%  \\ \hline
 \textbf{known divergence}       & 2.08\%   & 1.36\%  & 52\%   & 55.55\%  \\ \hline
 \textbf{known alignment}       & 0\%  & 0\%  & -  & -  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\end{tabular}}  & 20.95\%  & 24.71\%  & 93.03\%  & 94.76\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + head alignment\end{tabular}}  & 21.78\%  & 19.61\%  & 67.81\%  & 76.74\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}POS-equivalence\\ + head alignment\end{tabular}}  & 1.33\%   & 1.29\%  & 75\%   & 64.70\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + known alignment\end{tabular}}  & 0.25\%   & 0.45\%  & 66.66\%  & 83.33\%  \\ \hline
 \textbf{\begin{tabular}[c]{@{}l@{}}matching UD labels\\ + POS-equivalence\\ + head alignment\end{tabular}} & 10.85\%  & 12.69\%  & 95.38\%  & 90.41\%  \\ \hline
 \end{tabular}
 \caption[Criterion-specific statistics]{Criterion-specific statistics. The leftmost column specifies a criterion or combination of criteria. The central column indicates the percentage of alignments extracted because of each set of criteria, and the rightmost one specifies how many of them were marked as \texttt{+} or \texttt{-}.}
 \label{treas}
 \end{table}

\subsection{Working with raw text}
Of course, it is also interesting to compare the performance of the CE module on manually annotated data with the results obtained on the automatically parsed course plans corpora, summarized in Table \ref{tcereal}.
% Please add the following required packages to your document preamble:
% \usepackage[normalem]{ulem}
% \useunder{\uline}{\ul}{}
\begin{table}[H]
 \footnotesize
 \centering 
 \begin{tabular}{|l|l|l|}
 \hline
 \textbf{}   & \textbf{DMI (en-it, 798 sentences)} & \textbf{CSE (en-sv, 498 sentences)} \\ \hline
 \textbf{distinct alignments} & 352  & 529  \\ \hline
 \textbf{correct (+ and =)} & 243 (69.03\%) & 368 (69.56\%) \\ \hline
 \textbf{correct and useful (+)} & 229 (65.05\%) & 351 (66.35\%) \\ \hline
 \end{tabular}
 \caption[Performance of CE on barely sentence-aligned data]{Performance of CE on barely sentence-aligned data, parsed automatically. Two distinct corpora of different size have been used for the two different language pairs, so the two columns are not to be compared with each other, but rather with Table \ref{tbase}.}
 \label{tcereal}
\end{table}

@@

A Haskell script converts the CoNNL-U files that are fed to our CE module into the input format required by \texttt{fast\_align}, so that the exact same tokenization is used. \smallskip

@@

Because \texttt{fast\_align} is a purely statistical approach, we trained it not only on the same 100-sentence subset of PUD used for evaluating our proposed improved CA module against the baseline, but also on the full 1000-sentences dataset. 
In the latter experiment, the alignments obtained for the last 900 sentences were discarded before evaluating the results.
Due to the time required to manually annotate the numerous alignments extracted by \texttt{fast\_align}, this evaluation has been conducted only for the most challenging language pair, English-Italian. 
For a fairer comparison, since \texttt{fast\_align} works at the word level, only the one-to-one, one-to-many and many-to-one word alignments produced by the CE module were kept.

\begin{table}[H]
 \centering
 \small
 \begin{tabular}{|l|l|l|l|}
 \hline
  & \textbf{our system} & \textbf{\texttt{fast\_align} 100} & \textbf{\texttt{fast\_align} 1000}\\ \hline
  \textbf{distinct alignments} & 716 & 1440 & 1435 \\ \hline
  \textbf{correct} & 536 (74.86\%) & 410 (28.47\%) & 656 (45.71\%)\\ \hline
  \textbf{correct and useful} & 491 (68.57\%) & 371 (25.76\%) & 590 (41.11\%)\\ \hline
 \end{tabular}
 \caption[Comparison between the improved CE module and \texttt{fast\_align}]{Comparison between the improved CE module and \texttt{fast\_align} trained on the first 100 only (column 2) and on the whole PUD corpus (column 3).}
 \label{tfast}
\end{table}
